Timeliness of security updates for Android

### 2014-10-20 securityupdates.tex
Move some subsections around
Further changes based on Alastair's comments

\input{dastats}
\newcommand{\da}{Device Analyzer}
\input{avostats}
\newcommand{\avo}{\texttt{androidvulnerabilities.org}}
\newcommand{\percMarketShare}{XX\%~\cite{TODO}}
\newcommand{\daNumDevices}{\daNumOSDevices}
\newcommand{\daDeviceDays}{\daOSTotalDaysData}
% Num versions since \daStartDate
\input{countversions}
\newcommand{\apacheLibsNumVersions}{C}
\newcommand{\apacheLibsNumLibs}{E}

@@

\begin{abstract}
Android is the most popular smartphone platform and seeks to provide better security than popular desktop operating systems through increased compartmentalisation and a permission system for apps.
This security relies on the absence of root privilege exploits which allow apps to bypass protections.
Many such exploits have been found and fixed.
Our hypothesis is that these fixes do not promptly reach the users devices and so much of the time users are running versions of Android known to be vulnerable.
We used da data from over \daOSYearsOfData\ years and \daNumOSDevices\ devices and found that on average over \daMeanInsecurityPerc\ of devices were exposed to known root privilege vulnerabilities and only \daUpdatednessPerc\ of devices run the most recent version of Android.
There was also a period of several months when no devices ran secure versions of Android.
We compare different manufacturers and operators to find which are best at promptly distributing security updates and examine how to improve the situation. %TODO operators analysis
\end{abstract}

\section{Introduction}
Android has \percMarketShare\ of the smartphone market and while the core development is controlled by Google there are about \daNumManufacturers\footnote{\label{foot:dadata}We computed this from the da~\cite{Wagner2013} data, see \S\ref{sec:android_update_process}.} manufacturers which make devices which run Android.
Android receives regular updates  which add new features and fix bugs, including security vulnerabilities.
Many of these manufacturers customise the version of Android they ship and sometimes network operators (of which there are at least \daNumOperators\textsuperscript{\ref{foot:dadata}}) make further modifications.
Hence when Google produces an update to Android, the update may have to pass through the manufacturer and operator before reaching the user.
The manufacturer and operator have little financial incentive to perform this work as they have already sold the device.
Operators are cautious because if they ship a broken update then they will get many support requests which is expensive.
However there is ongoing legal action to force operators to ship updates for security vulnerabilities~\cite{Soghoian2013}.
Many phones are sold on long (12--24 month) contracts with monthly payments yet stop receiving updates before the end of the contract, while Windows XP could be purchased for a one off payment in October 2001 and received security updates until April 2014.
Corporate or public sector buyers are encouraged to purchase secure devices, but we have found little concrete guidance. For example, CESG, which advises the UK government on how to secure its computer systems, recommends picking Android devices from manufactures which are good at shipping security updates promptly~\cite{CESG2013} but it does not state which manufacturers are better.
We provide this information in~\S\ref{sec:security_scoring}.

@@

The proportion of Android devices in the da data which were running versions of Android known to have root vulnerabilities at the time is shown in Figure~\ref{fig:proportioninsecure}.
In \S\ref{sec:background} we will explain where these data come from and then in \S\ref{sec:results} we will show these data in more detail.

@@

\subsection{Root vulnerabilities}
We compiled a list of root vulnerabilities for Android containing information on when they were known about, what versions they affected and which versions fixed the problem.
We only looked for \emph{root equivalent} vulnerabilities which did not require USB debugging to exploit.
\emph{Root equivalent} means that it grants privileges equivalent in scope to root which can then either be used to gain root or provide sufficiently many permissions as to be as bad as root.
Some phones can be `rooted' by enabling USB debugging and using the special privileges of the adb shell to root the device but only \daAdbEnabledPerc\ of devices have USB debugging enabled.
This is not something that applications running on the phone can exploit to break compartmentalisation and so we do not include those vulnerabilities.
Unfortunately many published exploits use adb for convenience and so determining whether that is necessary to exploit the vulnerability can be difficult.
Some vulnerabilities which grant root are not traditional kernel vulnerabilities, for example the discovery of flaws in the verification of signatures on Android applications in February 2013~\cite{Forristal2013} meant that applications could pretend to be signed with system keys and hence gain root equivalent privileges.
On some versions of Android (below 4.1) they could use known system to root escalation mechanisms but on all versions they have a greatly increased attack area for further privilege escalation and also have the ability to control all user internet traffic (via VPNs), brick the phone, remove and install apps, steal user credentials, read the screen, make and receive calls.

@@

There was much manual effort required to track down details of these vulnerabilities as many of them are not in other databases such as the CVE database, the lack of a widely acknowledged unique identifier made identifying whether two reports referenced the same vulnerability difficult.
Previous work has assumed ``any security issue of relevance will eventually get a CVE number assigned"~\cite{Frei2010} which is currently not the case for Android with several root vulnerabilities not having CVE numbers.
For some of the vulnerabilities without CVE numbers we checked with Google whether they had one assigned and they confirmed that there was no number and that they did not intend to get one (though they did give us the Android bug number and tell us to use that).

@@

Prior work~\cite{Bilge2012} by Symantec on desktop malware has shown that after public disclosure exploitation rates increase by 5 orders of magnitude and so from the point of view of widespread danger, the period between public disclosure and the update reaching the user's device is the most critical.
However they also show that zero day vulnerabilities are typically used for 312 days before they are publicly disclosed, often to target particular organisations.
Hence when considering vulnerability from the point of view of an organisation which cares about Advanced Persistent Threats, such as those the CESG advice~\cite{CESG2013} is aimed at, the critical period is from when people could have started using the vulnerability to when the update hits the user device.
So for our vulnerability calculations we use the earliest date we can find recorded evidence that the vulnerability was known about, even if that knowledge might have been confined to a particular manufacturer or hobbyist.
We cannot know if someone reported the vulnerability to the manufacturer and we do know that various agencies are engaged in widespread monitoring of communications and compromise of civilian infrastructure~\cite{TODO} and so they probably know.
Hence we use dates such as: when CVEs were registered, when fixes were committed to source repositories, or when exploits were committed to repositories; as the date first known about.

@@

\subsection{Android update process}
\label{sec:android_update_process}
\begin{figure}
 \centering
 \def\svgwidth{\columnwidth}
 \import{figures/}{update_ecosystem.pdf_tex}
 \caption{Flow of updates between participants in the Android ecosystem.
 Numbers on edges indicate updates shipped between \daStartDate\ and \daEndDate, numbers in brackets represent number of such entities in our data.
 Dotted arrows indicate flows where we do not know how many updates are being produced.}
 \label{fig:update_ecosystem}
\end{figure}

A model of the Android update process is shown in Figure~\ref{fig:update_ecosystem}.
There are five entities or groups which contribute towards Android updates: the network operators, the device manufacturers, the hardware developers, Google and the upstream open source projects.
Android builds on various open source projects such as the Linux kernel, the OpenSSL and BouncyCastle cryptography libraries and various Apache Java libraries and so can include any of the versions of those projects, including those which fix security vulnerabilities in them.
It also incorporates various drivers for different bits of hardware.
The Android platform is then built on top of those with kernel and userspace components by Google. \todo{ work out the exact arrangements.}
Usually the code for each update is kept secret until after the update has been published though through the Open Handset Alliance some manufacturers see it in advance.
Then the Manufacturers take the update and customise it before passing it on to the operator.
The Operator may then make further customisations and do further testing before shipping the update to the Device.
Sometimes Manufactures ship direct to the user, sometimes the Manufacturer and Google are the same entity (at least for the purpose of a particular phone, such as with Nexus phones).
Sometimes Manufacturers incorporate upstream open source project releases directly, and sometimes incorrectly -- for example including a broken daily build of sqlite in their release of Android
%An analysis by Vidas et al.~\cite{Vidas2011} of the Android 2.1 to 2.2 update found that it took 11 months from when Google released 2.2 for the last device which they were investigating to get the update.
\todo{talk about update latency}

@@

\section{Experiments}
\label{sec:results}
In this section we present the results of our analysis showing \daMeanInsecurityPerc\ of devices to have root vulnerabilities on average, nexus devices to be better than others, \emph{\daSecScoreBestmanufacturer} to be the best manufacturer and \emph{\daSecScoreBestoperator} to be the best network operator. \todo{ summarise our results}
The \da\ data used here was collected between \daStartDate\ and \daEndDate.
The \avo\ data was collected between \avoStartDate\ and \avoEndDate.
\todo{comparison of nexus update rates with iOS}
We conducted four experiments examining the behaviour of the Android ecosystem as a whole (\S\ref{sec:exp:android_ecosystem}), updates reaching particular devices (\S\ref{sec:exp:device_updates}), comparisons between different manufacturers and operators (\S\ref{sec:exp:security_score}) and modeling the replacement by new Android versions in order to predict how long it will take for an update to be fixed (\S\ref{sec:exp:predicting_distributions}).
\todo{In an experiment you test a hypothesis, it needs to be clear what the hypothesis is and what the conclusion of the test is}
\subsection{Experiment 1: behaviour of the Android ecosystem}\label{sec:exp:android_ecosystem}
To investigate the behaviour of the Android ecosystem as a whole with respect to Android versions and vulnerabilities we used the OS version information from \da\ and the vulnerability data from \avo.
The proportion of devices in the \da\ data running different versions of Android each day is shown in Figure~\ref{fig:norm_os}.
It shows how old versions are gradually replaced by new ones, the long tail of devices which do not see updates to more recent versions and the fluctuation in the number of devices in the data.
We verify the validity of this data in \S\ref{sec:representative}.
\begin{figure}
 \centering
 \includegraphics[width=\columnwidth]{figures/da_norm_os}
 \caption{Android versions in \da\ data over time}
 \label{fig:norm_os}
\end{figure}

Figure~\ref{fig:vulnerabilities} shows which vulnerabilities devices are exposed to.
For each vulnerability it shows the proportion of devices exposed to that vulnerability and how that changes over time.
In July 2011 at the beginning of the \da\ data the \emph{exploid}\footnote{\url{http://androidvulnerabilities.org/vulnerabilities/exploid_udev}} and \emph{levitator}\footnote{\url{http://androidvulnerabilities.org/vulnerabilities/levitator}} vulnerabilities both affect most Android devices, slowly these are fixed as updates roll out and devices are replaced until in January 2013 a much smaller proportion of devices are affected by known vulnerabilities.
However when in February 2013 the first APK signing vulnerability was found which affected all previous versions of Android and even in October 2013 most devices were still vulnerable. \todo{put percentages on these things to quantify them}
\begin{figure}%[!b]
\centering
\includegraphics[width=\columnwidth]{figures/vulnerabilities}
\caption{Proportion of devices exposed to each vulnerability with time (1.0 is 100\%)}
\label{fig:vulnerabilities}
\end{figure}

The variation of the proportion of devices affected by a vulnerability with time tells us how bad a particular vulnerability affected the Android platform.
Figure~\ref{fig:nvulnerabilities_heat} breaks this down by vulnerability to show how the proportion of devices affected by different vulnerabilities varies.
In 2013 three vulnerabilities were found in the way which Android verified the signatures on APKs.
These allowed the creation of malicious APKs which appear to be signed as system APKs -- which have root equivalent privileges.
Figure~\ref{fig:nvulnerabilities_heat} shows how the the \emph{APK signing vulnerabilities} affected all devices and took months to get fixed for any device.
However what is perhaps more worrying is the long tail on the \emph{Gingerbreak}\footnote{\url{http://androidvulnerabilities.org/vulnerabilities/Gingerbreak}}, \emph{levitator} and \emph{exploid} vulnerabilities which are more dangerous root vulnerabilities (not requiring new APK installation) and which still affect a significant proportion of devices years later.

@@

\subsection{Update frequency}
\todo{ find out what the frequency of updates is}
\begin{figure}
 \includegraphics[width=\columnwidth]{figures/from_to_updates.pdf}
 \caption{Updates between different Android versions in the DA data}
 \label{fig:from_to_updates}
\end{figure}
\begin{figure}
 \includegraphics[width=\columnwidth]{figures/w_security_updates.pdf}
 \caption{Number of updates each week which fixed or may have fixed security vulnerabilities}
 \label{fig:weekly_security_updates}
\end{figure}
Figure~\ref{fig:from_to_updates} shows how devices upgrade between different versions of Android.
\da\ (DA) observes upgrade events and the figure shows which versions a device is upgrading from and to.
Mostly the dark cells are above the diagonal and are upgrades (\daNumUpdatesUpgrades).
While many upgrades are from one version to the next version there are also a fair number which skip a few versions.
Surprisingly there are also a small number of downgrade events (\daNumUpdatesDowngrades, \daPercUpdatesDowngrades) when older versions of Android are installed on to devices.
Possible reasons why users are downgrading are to free up space on their device, to make it easier to root or because a new version introduced bugs.

@@

The data appears to have an exponential decay as it tends to 1 but with a slow start as few devices are upgraded initially.
We approximate this with a a $e^{-x}$ component for the main behaviour and a which is offset to start later when uptake takes off.
The equation is then:
\begin{equation}
y =
 \begin{cases}
  1.0 & \text{if }x<\mathrm{undeployed}\\
  e^{-\mathrm{decay}(x-\mathrm{undeployed})} & \text{otherwise}
 \end{cases}
\end{equation}

Fitting that curve to the data using \texttt{scipy.optimize}, we get the parameters $\mathrm{undeployed} = \daOSCurveFitParamFirst ~\si{days}$, $\mathrm{decay} = \daOSCurveFitParamSecond ~\si{days^{-1}}$ for OS versions and $\mathrm{undeployed} = \daAPICurveFitParamFirst ~\si{days}$, $\mathrm{decay} = \daAPICurveFitParamSecond ~\si{days^{-1}}$ for API versions.
From this the number of days from release of a new version of Android until 50\% of devices are running that version or higher is \daOSCurveHalfDeployed\ and full deployment to \daFullDeployedAt\ of devices takes \daOSCurveFullDeployed\ days.
For API versions this is \daAPICurveHalfDeployed\ and \daAPICurveFullDeployed\ respectively.

@@

\section{Threats to validity}
\subsection{The \da\ data gives a conservative estimate of the Android version distribution}
\label{sec:representative}
\begin{figure}
 \centering
 \begin{subfigure}[b]{\columnwidth}
 \includegraphics[width=\columnwidth]{figures/googleplayapi}
 \caption{Google Play data on proportion of devices running different Android API versions}
 \label{fig:play_api}
\end{subfigure}
\begin{subfigure}[b]{\columnwidth}
 \includegraphics[width=\columnwidth]{figures/norm_api_gpcomp}
 \caption{\da\ data on proportion of devices running different Android API versions}
 \label{fig:da_api}
\end{subfigure}
\caption{Monthly Android API version data}
\end{figure}
\begin{figure}
 \centering
 \includegraphics[width=\columnwidth]{figures/api_gpcomp_rdiff}
 \caption{Difference between \da\ and Google Play data on the proportion of devices running different Android API versions}
 \label{fig:da_gp_comp_diff}
\end{figure}
The data from \da\ (DA) we used to investigate the proportion of devices exposed to different vulnerabilities is the OS version.
Unfortunately there is no authoritative source of OS version information and so we cannot check directly whether our data is representative.
However Google has published API version information every month since December 2009 and we have collated this information.\footnote{\url{https://www.cl.cam.ac.uk/~drt24/android/versiondistribution/}}
While API versions are too coarse grained to use for security update detection they are closely related to OS versions and so if the DA data on API versions is similar to the Google Data on API versions then the DA data on OS versions should be representative.
Figure~\ref{fig:play_api} shows the data from Google and Figure~\ref{fig:da_api} shows the data from \da\ and they appear similar.
Figure~\ref{fig:da_gp_comp_diff} shows the difference between Figures~\ref{fig:play_api} and \ref{fig:da_api}, normalising for days since the API version was released.
It shows that the \da\ data systematically overestimates the prevalence of new API versions and underestimates the prevalence of old API versions.
This means that the OS version information from \da\ is likely to be overestimating the prevalence of new OS versions and hence our results are a conservative estimate of the security of Android.
\todo{we want a statistical metric to claim this strongly with.}
This allows us to have confidence in the OS version information.

>>>>

\input{dastats}
\newcommand{\da}{Device Analyzer}
\newcommand{\dafoot}{\textsuperscript{\ref{foot:dadata}}}
\input{avostats}
\newcommand{\avo}{\texttt{androidvulnerabilities.org}}
\newcommand{\percMarketShare}{XX\%~\cite{TODO}}
\newcommand{\daNumDevices}{\daNumOSDevices}
\newcommand{\daDeviceDays}{\daOSTotalDaysData}
% Num versions since \daStartDate
\input{countversions}
\newcommand{\apacheLibsNumVersions}{C}
\newcommand{\apacheLibsNumLibs}{E}

@@

\begin{abstract}
Android is the most popular smartphone platform and seeks to provide better security than popular desktop operating systems through increased compartmentalisation and a permission system for apps.
This security relies on the absence of root privilege exploits which allow apps to bypass protections.
Many such exploits have been found and fixed.
Our hypothesis is that these fixes do not promptly reach the users devices and so much of the time users are running versions of Android known to be vulnerable.
We used \da\ data from over \daOSYearsOfData\ years and \daNumOSDevices\ devices and found that on average over \daMeanInsecurityPerc\ of devices were exposed to known root privilege vulnerabilities and only \daUpdatednessPerc\ of devices run the most recent version of Android.
There was also a period of several months when no devices ran secure versions of Android.
We compare different manufacturers and operators to find which are best at promptly distributing security updates and examine how to improve the situation. %TODO operators analysis
\end{abstract}

\section{Introduction}
Android has \percMarketShare\ of the smartphone market and while the core development is controlled by Google there are about \daNumManufacturers\footnote{\label{foot:dadata}We computed this from the \da~\cite{Wagner2013} data, see \S\ref{sec:android_update_process}.} manufacturers which make devices which run Android.
Android receives regular updates  which add new features and fix bugs, including security vulnerabilities.
Many of these manufacturers customise the version of Android they ship and sometimes network operators (of which there are at least \daNumOperators\dafoot) make further modifications.
Hence when Google produces an update to Android, the update may have to pass through the manufacturer and operator before reaching the user.
The manufacturer and operator have little financial incentive to perform this work as they have already sold the device.
Operators are cautious because if they ship a broken update then they will get many support requests which is expensive.
However there is ongoing legal action to force operators to ship updates for security vulnerabilities~\cite{Soghoian2013}.
Many phones are sold on long (12--24 month) contracts with monthly payments yet stop receiving updates before the end of the contract, while Windows XP could be purchased for a one off payment in October 2001 and received security updates until April 2014.
Corporate or public sector buyers are encouraged to purchase secure devices, but we have found little concrete guidance. For example, CESG, which advises the UK government on how to secure its computer systems, recommends picking Android devices from manufactures which are good at shipping security updates promptly~\cite{CESG2013} but it does not state which manufacturers are better.
We provide this information in~\S\ref{sec:security_scoring}.

@@

The proportion of Android devices in the \da\ data which were running versions of Android known to have root vulnerabilities at the time is shown in Figure~\ref{fig:proportioninsecure}.
In \S\ref{sec:background} we will explain where these data come from and then in \S\ref{sec:results} we will show these data in more detail.

@@

\subsection{Root vulnerabilities}
We compiled a list of root vulnerabilities for Android containing information on when they were known about, what versions they affected and which versions fixed the problem.
We only looked for \emph{root equivalent} vulnerabilities which did not require USB debugging to exploit.
\emph{Root equivalent} means that it grants privileges equivalent in scope to root which can then either be used to gain root or provide sufficiently many permissions as to be as bad as root.\todo{refine this definition}
Some phones can be `rooted' by enabling USB debugging and using the special privileges of the adb shell to root the device but only \daAdbEnabledPerc\dafoot\ of devices have USB debugging enabled.
This is not something that applications running on the phone can exploit to break compartmentalisation and so we do not include those vulnerabilities.
Unfortunately many published exploits use adb for convenience and so determining whether that is necessary to exploit the vulnerability can be difficult.
Some vulnerabilities which grant root are not traditional kernel vulnerabilities, for example the discovery of flaws in the verification of signatures on Android applications in February 2013~\cite{Forristal2013} meant that applications could pretend to be signed with system keys and hence gain root equivalent privileges.
On some versions of Android (below 4.1) they could use known system to root escalation mechanisms but on all versions they have a greatly increased attack area for further privilege escalation and also have the ability to control all user internet traffic (via VPNs), brick the phone, remove and install apps, steal user credentials, read the screen and make as well as receive calls.

@@

Tracking vulnerabilities is a manual task as they are not all in other databases such as the CVE database and the lack of a widely acknowledged unique identifier made identifying whether two reports referenced the same vulnerability difficult.
Previous work has assumed ``any security issue of relevance will eventually get a CVE number assigned"~\cite{Frei2010} which is currently not the case for Android root vulnerabilities.
For some of the vulnerabilities without CVE numbers, Google confirmed that there was no CVE number and that they did not intend to get one, instead providing an Android bug number.

@@

Prior work~\cite{Bilge2012} by Symantec on desktop malware has shown that after public disclosure exploitation rates increase by 5 orders of magnitude and so from the point of view of widespread danger, the period between public disclosure and the update reaching the user's device is the most critical.
However they also show that zero day vulnerabilities are typically used for 312 days before they are publicly disclosed, often to target particular organisations.
Hence when considering vulnerability from the point of view of an organisation which cares about Advanced Persistent Threats, such as those the CESG advice~\cite{CESG2013} is aimed at, the critical period is from when people could have started using the vulnerability to when the update hits the user device.
So for our vulnerability calculations we use the earliest date we can find recorded evidence that the vulnerability was known about, even if that knowledge might have been confined to a particular manufacturer or hobbyist.
We cannot know if someone reported the vulnerability to the manufacturer and we do know that various agencies are engaged in widespread monitoring of communications and compromise of civilian infrastructure~\cite{TODO} and so they probably know.
Hence we use dates such as: when CVEs were registered, when fixes were committed to source repositories, or when exploits were committed to repositories; as the date first known about.\todo{We need to be more precise here}

@@

\subsection{Android update process}
\label{sec:android_update_process}
\begin{figure}
 \centering
 \def\svgwidth{\columnwidth}
 \import{figures/}{update_ecosystem.pdf_tex}
 \caption{Flow of updates between participants in the Android ecosystem.
 Numbers on edges indicate updates shipped between \daStartDate\ and \daEndDate, numbers in brackets represent number of such entities in our data.
 Dotted arrows indicate flows where we do not know how many updates are being produced as we can't measure those flows directly as they are not public.\todo{Update this diagram}}
 \label{fig:update_ecosystem}
\end{figure}

A model of the Android update process is shown in Figure~\ref{fig:update_ecosystem}.
There are five entities or groups which contribute towards Android updates: the network operators, the device manufacturers, the hardware developers, Google and the upstream open source projects.
Android builds on various open source projects such as the Linux kernel, the OpenSSL and BouncyCastle cryptography libraries and various Apache Java libraries and so can include any of the versions of those projects, including those which fix security vulnerabilities in them.
It also incorporates various drivers for different bits of hardware.
The Android platform is then built on top of those with kernel and userspace components by Google. \todo{ work out the exact arrangements.}
Usually the code for each update is kept secret\todo{Can we quantify this keeping the code secret? Is it worth it?} until after the update has been published though through the Open Handset Alliance.
Manufacturers in the  alliance receive advanced copies in order to prepare handsets.
Then the Manufacturers take the update and customise it before passing it on to the operator.
The Operator may then make further customisations and do further testing before shipping the update to the Device.
Sometimes Manufactures ship direct to the user, sometimes the Manufacturer and Google are the same entity (at least for the purpose of a particular phone, such as with Nexus phones).
Sometimes Manufacturers incorporate upstream open source project releases directly, and sometimes incorrectly -- for example including a broken daily build of sqlite in their release of Android
%An analysis by Vidas et al.~\cite{Vidas2011} of the Android 2.1 to 2.2 update found that it took 11 months from when Google released 2.2 for the last device which they were investigating to get the update.
\todo{talk about update latency, quantify this in the text as per the figure, giving more explanation, use statistics from samsung-updates.com -> how many binaries are there per device? Do we know the variation between network operators?}

@@

\section{Experiments}
\label{sec:results}
In this section we present the results of our analysis showing \daMeanInsecurityPerc\ of devices to have root vulnerabilities on average. Nexus devices are better than others, \emph{\daSecScoreBestmanufacturer} to be the best manufacturer and \emph{\daSecScoreBestoperator} to be the best network operator. \todo{ summarise our results}
The \da\ data used here was collected between \daStartDate\ and \daEndDate.
The \avo\ data was collected between \avoStartDate\ and \avoEndDate.
\todo{comparison of nexus update rates with iOS}
We conducted four experiments examining the behaviour of the Android ecosystem as a whole (\S\ref{sec:exp:android_ecosystem}), updates reaching particular devices (\S\ref{sec:exp:device_updates}), comparisons between different manufacturers and operators (\S\ref{sec:exp:security_score}) and modeling the replacement by new Android versions in order to predict how long it will take for an update to be fixed (\S\ref{sec:exp:predicting_distributions}).
\todo{In an experiment you test a hypothesis, it needs to be clear what the hypothesis is and what the conclusion of the test is}
\subsection{Experiment 1: behaviour of the Android ecosystem}\label{sec:exp:android_ecosystem}
To investigate the behaviour of the Android ecosystem as a whole with respect to Android versions and vulnerabilities we used the OS version information from \da\ and the vulnerability data from \avo.
The proportion of devices in the \da\ data running different versions of Android each day is shown in Figure~\ref{fig:norm_os}.
It shows how old versions are gradually replaced by new ones, the long tail of devices which do not see updates to more recent versions and the fluctuation in the number of devices in the data.
We verify the validity of this data in \S\ref{sec:representative}.
\begin{figure}
 \centering
 \includegraphics[width=\columnwidth]{figures/da_norm_os}
 \caption{Android versions in \da\ data over time}
 \label{fig:norm_os}
\end{figure}

Figure~\ref{fig:vulnerabilities} shows which vulnerabilities devices are exposed to.
For each vulnerability it shows the proportion of devices exposed to that vulnerability and how that changes over time.
In July 2011 at the beginning of the \da\ data the \emph{exploid}\footnote{\url{http://androidvulnerabilities.org/vulnerabilities/exploid_udev}} and \emph{levitator}\footnote{\url{http://androidvulnerabilities.org/vulnerabilities/levitator}} vulnerabilities both affect most Android devices, slowly these are fixed as updates roll out and devices are replaced until in January 2013 a much smaller proportion of devices are affected by known vulnerabilities.\todo{Quantify whether it is replacement of updates which causes the fixes}
However when in February 2013 the first APK signing vulnerability was found which affected all previous versions of Android and even in October 2013 most devices were still vulnerable. \todo{put percentages on these things to quantify them}
\begin{figure}%[!b]
\centering
\includegraphics[width=\columnwidth]{figures/vulnerabilities}
\caption{Proportion of devices exposed to each vulnerability with time (1.0 is 100\%)}
\label{fig:vulnerabilities}
\end{figure}

The variation of the proportion of devices affected by a vulnerability with time tells us how bad a particular vulnerability affected the Android platform.
Figure~\ref{fig:nvulnerabilities_heat} breaks this down by vulnerability to show how the proportion of devices affected by different vulnerabilities varies.
In 2013 three vulnerabilities were found in the way which Android verified the signatures on APKs.
These allowed the creation of malicious APKs which appear to be signed as system APKs -- which have root equivalent privileges.
Figure~\ref{fig:nvulnerabilities_heat} shows how the the \emph{APK signing vulnerabilities} affected all devices and took months to get fixed for any device.
However what is perhaps more worrying is the long tail on the \emph{Gingerbreak}\footnote{\url{http://androidvulnerabilities.org/vulnerabilities/Gingerbreak}}, \emph{levitator} and \emph{exploid} vulnerabilities which are more dangerous root vulnerabilities (not requiring new APK installation) and which still affect a significant proportion of devices years later.
\todo{break things down by attack vector?}

@@

\subsubsection{Update frequency}
\todo{ find out what the frequency of updates is}
\begin{figure}
 \includegraphics[width=\columnwidth]{figures/from_to_updates.pdf}
 \caption{Updates between different Android versions in the \da data}
 \label{fig:from_to_updates}
\end{figure}
\begin{figure}
 \includegraphics[width=\columnwidth]{figures/w_security_updates.pdf}
 \caption{Number of updates each week which fixed or may have fixed security vulnerabilities}
 \label{fig:weekly_security_updates}
\end{figure}
Figure~\ref{fig:from_to_updates} shows how devices upgrade between different versions of Android.
\da\ (DA) observes upgrade events and the figure shows which versions a device is upgrading from and to.
Mostly the dark cells are above the diagonal and are upgrades (\daNumUpdatesUpgrades).
While many upgrades are from one version to the next version there are also a fair number which skip a few versions.
Surprisingly there are also a small number of downgrade events (\daNumUpdatesDowngrades, \daPercUpdatesDowngrades) when older versions of Android are installed on to devices.
Possible reasons why users are downgrading are to free up space on their device, to make it easier to root or because a new version introduced bugs.

@@

The data appears to have an exponential decay as it tends to 1 but with a slow start as few devices are upgraded initially.
We approximate this with a $e^{-x}$ for the main behaviour and a which is offset to start later when uptake takes off.
The equation is then:
\begin{equation}
y =
 \begin{cases}
  1.0 & \text{if }x<\mathrm{undeployed}\\
  e^{-\mathrm{decay}(x-\mathrm{undeployed})} & \text{otherwise}
 \end{cases}
\end{equation}

Fitting that curve to the data using \texttt{scipy.optimize}, we get a RMSE of \daOSCurveFitRMSE\ and parameters $\mathrm{undeployed} = \daOSCurveFitParamFirst ~\si{days}$, $\mathrm{decay} = \daOSCurveFitParamSecond ~\si{days^{-1}}$ for OS versions and RMSE of \daAPICurveFitRMSE\, $\mathrm{undeployed} = \daAPICurveFitParamFirst ~\si{days}$, $\mathrm{decay} = \daAPICurveFitParamSecond ~\si{days^{-1}}$ for API versions.
From this the number of days from release of a new version of Android until 50\% of devices are running that version or higher is \daOSCurveHalfDeployed\ and full deployment to \daFullDeployedAt\ of devices takes \daOSCurveFullDeployed\ days.
For API versions this is \daAPICurveHalfDeployed\ and \daAPICurveFullDeployed\ respectively.

@@

\section{Threats to validity}
\subsection{The \da\ data gives a conservative estimate of the Android version distribution}
\label{sec:representative}
\begin{figure}
 \centering
 \begin{subfigure}[b]{\columnwidth}
 \includegraphics[width=\columnwidth]{figures/googleplayapi}
 \caption{Google Play data on proportion of devices running different Android API versions}
 \label{fig:play_api}
\end{subfigure}
\begin{subfigure}[b]{\columnwidth}
 \includegraphics[width=\columnwidth]{figures/norm_api_gpcomp}
 \caption{\da\ data on proportion of devices running different Android API versions}
 \label{fig:da_api}
\end{subfigure}
\caption{Monthly Android API version data}
\end{figure}
\begin{figure}
 \centering
 \includegraphics[width=\columnwidth]{figures/api_gpcomp_rdiff}
 \caption{Difference between \da\ and Google Play data on the proportion of devices running different Android API versions}
 \label{fig:da_gp_comp_diff}
\end{figure}
The data from \da\ (DA) we used to investigate the proportion of devices exposed to different vulnerabilities is the OS version.
Unfortunately there is no authoritative source of OS version information and so we cannot check directly whether our data is representative.
However Google has published API version information every month since December 2009 and we have collated this information.\footnote{\url{https://www.cl.cam.ac.uk/~drt24/android/versiondistribution/}}
While API versions are too coarse grained to use for security update detection they are closely related to OS versions and so if the \da\ data on API versions is similar to the Google Data on API versions then the \da\ data on OS versions should be representative.
Figure~\ref{fig:play_api} shows the data from Google and Figure~\ref{fig:da_api} shows the data from \da\ and they appear similar.
Figure~\ref{fig:da_gp_comp_diff} shows the difference between Figures~\ref{fig:play_api} and \ref{fig:da_api}, normalising for days since the API version was released.
It shows that the \da\ data systematically overestimates the prevalence of new API versions and underestimates the prevalence of old API versions.
This means that the OS version information from \da\ is likely to be overestimating the prevalence of new OS versions and hence our results are a conservative estimate of the security of Android.
\todo{we want a statistical metric to claim this strongly with.}
This allows us to have confidence in the OS version information.
